---
title: "Data Cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(dplyr)
library(tidyr)
```

## Data Sources

Read in the data from the [KNB](https://knb.ecoinformatics.org/#view/df35b.304.2). 

```{r}
catch_original <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

glimpse(catch_original)
```

## Clean and Reshape Data

* remove unnecessary columns

```{r}
catch_data <- catch_original %>% 
  # select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) Can list what you want to keep
  select(-All, -notesRegCode) #or what you want to get rid of with a minus symbol
```

* check column types

```{r}
#catch_clean <- catch_data %>% 
  #mutate(Chinook = as.numeric()) #as.numeric will try to coerce the column Chinook to numeric form (it was character). This code did not run, so the next code chunk will not run because catch_clean was not created here.

```

```{r}
#i <- which(is.na(catch_clean$Chinook))
#i

#catch_original[i,] #in one row, the entry for Chinook is "I" not "1", go back to original data sheets to see if it is an error or a footnote
```

```{r}
catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>%  #if Chinook is equal to I, make it one, otherwise, leave it alone
  mutate(Chinook = as.integer(Chinook)) #switched to as.integer to match the other column types in the dataset

glimpse(catch_clean)  
```


* reshape data

```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year), names_to = "species", values_to = "catch")

head(catch_long)
```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Region, values_from = catch)

head(catch_wide)
```

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch) %>%  #rename the catch column to include units
  mutate(catch = catch_thousands * 1000) #add a new catch column that is actually the number of fish

head(catch_long)
```

* group and summarize data

```{r}
mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)
```

```{r}
mean_year <- catch_long %>% 
  group_by(Year) %>% 
  summarize(catch_mean = mean(catch))

head(mean_year)
```

Using multiple grouping variables.  

```{r}
mean_year_species <- catch_long %>% 
  group_by(Year, species) %>% 
  summarize(catch_mean = mean(catch))

head(mean_year_species)
```

Using multiple grouping variables and multiple statistics.  

```{r}
mean_region <- catch_long %>% 
  group_by(Region, Year) %>% 
  summarize(catch_mean = mean(catch), num_obs = n(), .groups = "drop") #can drop the groups to prevent grouping errors later

head(mean_region)
```


## Join to Regions Definitions 

```{r}
region_defs <- read.csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1") %>% 
    select(code, mgmtArea)

head(region_defs)
```


```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

head(catch_joined)
```

Mattea is creating a git conflict here. Hi Mattea, this is Cat! Mattea is added second conflict here.

